State of the Practice for Lattice Boltzmann Method Software

### 2022-03-06 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
reviewed abstract and section 2 - methodology

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be instantiated
for a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step of the steps in
the methodology.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case effort includes the time spent finding and understanding the
	installation instructions, the person-time and resources spent performing
	the installation procedure, and the absence or ease of overcoming system
	compatibility issues. An installability score improves if the software has a
	means to validate the installation and instructions for uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
	the specification is available. Research software is unlikely to have an
	explicit specification, since this is not common practice in the field. As a
	consequence, the correctness of software cannot often be measured directly.
	We assess correctness indirectly by looking for the following: availability
	of a requirements specification, reference to domain theory, and explicit
	use of tools and/or techniques for building confidence, such as
	documentation generators and software analysis tools.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites.

	\item[Modifiability] Modifiability refers to the ease with which stable
	changes can be made to a system and the flexibility of the system to adopt
	such changes \citep{8016712}. We did not directly measure modifiability.
	Instead, developers were asked in interviews if they considered the ease of
	future changes when developing the software packages, specifically changes
	to the structure of the system, modules and code blocks. A follow up
	question is asked if any measures have been taken.

@@

To reduce the length of the list of software packages and to
answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45.  The following
filters were applied in the priority order listed.

\begin{enumerate}
	\item Scope: Software is removed by narrowing what functionality is
	considered to be within the scope of the domain.
	\item Usage: Software packages were eliminated if their installation
	procedure was missing or not clear and easy to follow.
	\item Age: The older software packages (age being measured by the last date
	when a change was made) were eliminated, except in the cases where an older
	software package appears to be highly recommended and currently in use. 
\end{enumerate}

@@

Several of the qualities use the word ``surface''.  This is to highlight that,
for these qualities in particular, the best that we can do is a shallow measure.
For instance, we do not conduct experiments to measure usability. Instead, we
are looking for an indication that usability was considered by the developers by
looking for cues in the documentation, such as getting started instructions, a
user manual or documentation of expected user characteristics.

@@

\subsection{Analytical Hierarchy Process} \label{AHP}

Once we have measured each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}.  To do this, we used the Analytical Hierarchy Process
(AHP), a decision-making technique that is used to compare multiple options by
multiple criteria. AHP was used for comparing and ranking the LBM software
packages using the overall impression quality scores that were gathered in the
measurement template.  AHP performs a pairwise analysis using a matrix and
generates an overall score as well as individual quality scores for each
software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
software based on quality measures.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interviewed developers
using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
are about specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often bring up exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meeting were held using on-line
meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

>>>>

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be applied
to a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step in the methodology.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case, effort includes the time spent finding and understanding the
	installation instructions, the person-time and resources spent performing
	the installation procedure, and the absence or ease of overcoming system
	compatibility issues. An installability score improves if the software has a
	means to validate the installation and instructions for uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
	the specification is available. Research software is unlikely to have an
	explicit specification, since this is not common practice in the field. As a
	consequence, the correctness of software often cannot be measured directly.
	We assess correctness indirectly by looking for the following: availability
	of a requirements specification, reference to domain theory, and explicit
	use of tools and/or techniques for building confidence, such as
	documentation generators and software analysis tools.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work,
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites.

	\item[Modifiability] Modifiability refers to the ease with which stable
	changes can be made to a system and the flexibility of the system to adopt
	such changes \citep{8016712}. We did not directly measure modifiability.
	Instead, developers were asked in interviews if they considered the ease of
	future changes when developing the software packages, specifically changes
	to the structure of the system, modules and code blocks. A follow up
	question asked if any measures have been taken to ensure the ease of future changes.

@@

To reduce the length of the list of software packages and to
answer~\rqref{RQ_WhatProjects}, we filter the initial list of 45 packages.  The following filters were applied in the priority order listed.

\begin{enumerate}
	\item Scope: Software is removed by narrowing what functionality is
	considered to be within the scope of the domain.
	\item Usage: Software packages are eliminated if their installation
	procedure was missing or not clear and easy to follow.
	\item Age: The older software packages (age being measured by the last date
	when a change was made) are eliminated, except in the cases where an older
	software package appears to be highly recommended and currently in use. 
\end{enumerate}

@@

Several of the qualities use the word ``surface''.  This is to highlight that,
for these qualities in particular, the best that we can do is a shallow measure.
For instance, we do not conduct experiments to measure usability. Instead, we
are looking for an indication that usability was considered by the developers by
looking for cues in the documentation, such as the presence of getting started instructions, a user manual or documentation of expected user characteristics.

@@

\subsection{Analytic Hierarchy Process} \label{AHP}

Once we have measured each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}.  To do this, we used the Analytic Hierarchy Process
(AHP), a decision-making technique that is used to compare multiple options by
multiple criteria. AHP was used for comparing and ranking the LBM software
packages using the overall impression quality scores that were gathered in the
measurement template.  AHP performs a pairwise analysis using a matrix and
generates an overall score as well as individual quality scores for each
software package. \citep{SmithEtAl2016} shows how AHP is applied to ranking
software based on quality measures.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interviewed developers
using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
focus on specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often bring up exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meetings were held using on-line meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.
